---
title: "Intro to Water Quality Analysis (R you ready!)"
author: "Matthew Ross"
date: "`r format(Sys.time(), '%B %d, %Y')`"
output:
  ioslides_presentation:
    incremental: true
    widescreen: true
    smaller: true
    transition: 0.01
editor_options: 
  chunk_output_type: console
---

<style>
slides > slide {
  overflow-x: auto !important;
  overflow-y: auto !important;
}
</style>


## Why data analysis in a water quality lab? 

- Big Data, everywhere all the time
- Sensors
- Jobs

## Why R for data analysis?

- Free
- Dynamic
- Powerful
- Welcoming
- Most importantly... I use it!

## I've never used R and you're scaring me
- Not a problem
- We'll provide all the support you need to incorporate it into your work for this 
class

## Some core basics {.build}

- R is an object oriented language
- This means specific data types can do specific things.

```{r numeric vector}
#Make a vector 
this.is.a.numeric.vector = c(1,2)
#Print the vector
print(this.is.a.numeric.vector)
#Look at the structure
str(this.is.a.numeric.vector)
#Depending on object type (numeric vector here) # We can do specific things
this.is.a.numeric.vector*2
```

## Errors {.build}

- When you try to do something to an object that it doesn't want to do, you get an error
```{r character vector,error=T}

#Let's make a character vector by quoting our numbers
this.is.not.a.numeric.vector=c('1','2')
#Print the vector
print(this.is.not.a.numeric.vector)
#Look at its structure
str(this.is.not.a.numeric.vector)
#Try to multiply a character vector by 2
this.is.not.a.numeric.vector*2
```

## Data frames {.build}
- If you use excel, data frames are like excel sheets
- They have labeled columns and rows of data
- They are the primary data object you will work with

```{r data frame}
#Bind the two vectors from above
this.is.a.data.frame = data.frame(numbers=this.is.a.numeric.vector,
                                  characters=this.is.not.a.numeric.vector)
##
print(this.is.a.data.frame)

```


## 8 years later...
- R is very powerful and should eventually make your work easier, more beautiful,
and faster. 
- This will take some time, but the following example should make it clear
why it's a valuable tool to at least be familiar with

## Install and use libraries

```{r setup, include=T,warning=F,message=F}
library(tidyverse)
library(dataRetrieval)
library(sf) #Spatial package
library(mapview) # interactive mapper
library(xts) # Time series package
library(dygraphs) #Time series plotting package
knitr::opts_chunk$set(echo = TRUE)
knitr::opts_knit$set(root.dir='..')
```



## Download and plot USGS gauging stations with discharge data {.build}
```{r}
#Download site information
usgs.sites <- whatNWISsites(stateCd='CO',parameterCd='00060')

#Print first six rows of data
head(usgs.sites)
#Print the data structure
str(usgs.sites)

#check names of dataset
names(usgs.sites)
#Plot the data in a map
sites.map <- st_as_sf(usgs.sites,
                      coords=c('dec_long_va','dec_lat_va'),
                      crs=4326)

#Examine structure of sites map
str(sites.map)

```

## Dynamically plot USGS site locations
```{r,out.width='100%'}
mapviewOptions(leafletWidth='100%')
mapview(sites.map)

```



## Download fort collins Q data for 2015-2019 {.build}
```{r}
#Site codes from map
site_cd_foco <- c('06752260','06752280')


usgs.site.data <- readNWISdata(sites=site_cd_foco,
                               service='dv',
                               parameterCd='00060',
                               startDate='2015-01-01',
                               endDate='2019-01-15')

```

## Change data structure {.build}
```{r}
#Look at names of data frame
names(usgs.site.data)

#simplify data 
foco.data <- usgs.site.data %>%
  #Rename site codes into site names
  mutate(site_name=ifelse(site_no=='06752260','Lincoln','ELC')) %>%
  #Rename and remove useless columns
  dplyr::select(site_name,date=dateTime,q.cfs=X_00060_00003) %>%
  #Filter out bad data
  filter(q.cfs > 0)

#Look at new data frame
head(foco.data)
```

## Alter data for plotting {.build=T}
```{r}
## Alter data structure to match plotting library
foco.wide <- spread(foco.data,
                    key=site_name,
                    value=q.cfs)

#Convert to a timeseries object. 
foco.ts <- xts(x=foco.wide %>%
                 select(-date),
               order.by=foco.wide$date)

head(foco.ts)
```


## Dynamic plot of Q data
```{r}
dygraph(foco.ts)
```

## Summary
- R is awesome
- It will be used in this class
- We'll help you use it
- It's mostly copying and pasting the internet anyways (with attribution please!)